Analyzing TTC Delay Data to Improve Future Efficiency

### 2025-06-20 outputs/paper.qmd
fixed issue -- papr.qmd in python now

```{r}
#| echo: false
#| message: false
#| warning: false

#### Workspace setup ####
library(janitor)
library(knitr)
library(opendatatoronto)
library(tidyverse)

@@

```{r}
#| echo: false
#| message: false
#| warning: false

# Read in cleaned data
bus_delay_data <- 
  read_csv(
     file = "data/cleaned_bus_delay_statistics.csv",
     show_col_types = FALSE
  )

subway_delay_data <-
  read_csv(
     file = "data/cleaned_subway_delay_statistics.csv",
     show_col_types = FALSE
  )
```

@@

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-bus_delay_data
#| tbl-cap: Sample of Bus Delay Statistics

# Display a sample of the bus dataset in a table
bus_delay_data |>
  head(5) |>
  kable(
    col.names = c("Date", "Time", "Day", "Incident", "Min Delay", "Min Gap"),
    booktabs = TRUE
  )
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-subway_delay_data
#| tbl-cap: Sample of Subway Delay Statistics

# Display a sample of the subway dataset in a table
subway_delay_data |>
  head(5) |>
  kable(
    col.names = c("Date", "Time", "Day", "Incident", "Min Delay", "Min Gap", "Line"),
    booktabs = TRUE
  )

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-vstime1
#| fig-cap: "The amount of delays with respect to the time of day"
#| fig-width: 4
#| fig-height: 3

# Display the total amount of delays (on any date) for
# a particular time

bus_delay_data |>
  ggplot(aes(x = time)) +
  geom_histogram(fill = "blue", alpha = 0.75) +
  scale_x_time(breaks = scales::pretty_breaks(n = 12)) +
  xlab("Time") +
  ylab("Count") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-vstime2
#| fig-cap: "The amount of delays with respect to the time of day"
#| fig-width: 4
#| fig-height: 3

# Display the total amount of delays (on any date) for
# a particular time

subway_delay_data |>
  ggplot(aes(x=time)) +
  geom_histogram(fill = "red", alpha = 0.75) +
  scale_x_time(breaks = scales::pretty_breaks(n = 12)) +
  xlab("Time") +
  ylab("Count") +
  theme_light() + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-avg_num_vsday
#| fig-cap: "The average number of delays occurred on a given day of the week"
#| fig-subcap: ["Bus Delays", "Subway Delays"]
#| layout-ncol: 2

# Load in summarized data
avg_num_bus_delays_by_day <- read_csv(
   file = "data/summaries/avg_num_bus_delays_by_day.csv",
   show_col_types = FALSE
)

avg_num_bus_delays_by_day$day <- 
  factor(avg_num_bus_delays_by_day$day, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))

avg_num_bus_delays_by_day |>
  ggplot(aes(x = day, y = mean_num_delays, fill = day)) +
  geom_bar(stat = 'identity') +
  xlab("Day of the Week") +
  ylab("Mean Number of Delays") + 
  theme_light() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

avg_num_subway_delays_by_day <- read_csv(
   file = "data/summaries/avg_num_subway_delays_by_day.csv",
   show_col_types = FALSE
)

avg_num_subway_delays_by_day$day <- 
  factor(avg_num_subway_delays_by_day$day, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))

avg_num_subway_delays_by_day |>
  ggplot(aes(x = day, y = mean_num_delays, fill = day)) +
  geom_bar(stat = 'identity') +
  xlab("Day of the Week") +
  ylab("Mean Number of Delays") + 
  theme_light() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: tbl-totalbus
#| tbl-cap: "The total and average time in minutes wasted on delays"

total_bus_delay_time_by_date <- 
  read_csv(
   file = "data/summaries/total_bus_delay_time_by_date.csv",
   show_col_types = FALSE
)

# Take 5 random rows from the summarized data 
set.seed(302)
total_bus_delay_time_by_date[sample(nrow(total_bus_delay_time_by_date), 5), ] |>
  kable(
    col.names = c("Date", "Day", "Total Delay Time", "Delay Count", "Average Delay Time"),
    booktabs = TRUE
  )

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-fullyear
#| fig-cap: "Bus Total and Average Delays in Time (Minutes)"
#| fig-subcap: ["Total Delay Time", "Average Delay Time"]
#| layout-ncol: 2

total_bus_delay_time_by_date |>
  ggplot(aes(x = date, y = total_delay_time)) + 
  geom_point() +
  geom_smooth(method = "loess") +
  theme_light() +
  xlab("Date") +
  ylab("Total Delay Time (Minutes)")

total_bus_delay_time_by_date |>
  ggplot(aes(x = date, y = mean_delay_time)) + 
  geom_point() +
  geom_smooth(method = "loess") +
  theme_light() +
  xlab("Date") +
  ylab("Total Delay Time (Minutes)")

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-vsincident1
#| fig-cap: "The average number of delays occurred due to incident types"

# Load summarized data about number of delays vs incidents

bus_delays_by_incident <- read_csv(
   file = "data/summaries/total_num_bus_delays_by_incident.csv",
   show_col_types = FALSE
) 
bus_delays_by_incident |> 
  ggplot(aes(x=incident, y=n, fill=incident)) +
  geom_bar(stat = "identity") +
  xlab("Incident Type") +
  ylab("Delay Count") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  guides(fill=guide_legend(title="Incident Type"))

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-vsincident2
#| fig-cap: "The average number of delays occurred due to incident types"

subway_delays_by_incident <- read_csv(
   file = "data/summaries/total_num_subway_delays_by_incident.csv",
   show_col_types = FALSE
) 
subway_delays_by_incident |> 
  ggplot(aes(x = incident, y = n, fill = incident)) +
  geom_bar(stat = "identity") +
  xlab("Incident Type") +
  ylab("Delay Count") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  guides(fill=guide_legend(title="Incident Type"))

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-vsline
#| fig-cap: "The number of delays on each TTC Line"

subway_delays_by_line <- read_csv(
   file = "data/summaries/total_num_subway_delays_by_line.csv",
   show_col_types = FALSE
)

subway_delays_by_line |> 
  ggplot(aes(x = line, y = n, fill = line)) +
  geom_bar(stat = "identity") + 
  theme_light() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  xlab("TTC Line") +
  ylab("Number of Delays")

>>>>

```{python}
#| echo: false
#| message: false
#| warning: false

#### Workspace setup ####
import matplotlib.pyplot as plt
import pandas as pd

@@

```{python}
#| echo: false
#| message: false
#| warning: false

# Read bus delay data
bus_delay_data = pd.read_csv("data/cleaned_bus_delay_statistics.csv")

# Read subway delay data
subway_delay_data = pd.read_csv("data/cleaned_subway_delay_statistics.csv")
```

@@

```{python}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-bus_delay_data
#| tbl-cap: Sample of Bus Delay Statistics

# Display a sample of the bus dataset in a table
sample_bus_delay_data = bus_delay_data.head(5)
sample_bus_delay_data.columns = ["Date", "Time", "Day", "Incident", "Min Delay", "Min Gap"]
sample_bus_delay_data
```

```{python}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-subway_delay_data
#| tbl-cap: Sample of Subway Delay Statistics

# Display a sample of the subway dataset in a table
sample_subway_delay_data = subway_delay_data.head(5)
sample_subway_delay_data.columns = ["Date", "Time", "Day", "Incident", "Min Delay", "Min Gap", "Line"]
sample_subway_delay_data

@@

```{python}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-vstime1
#| fig-cap: "The amount of delays with respect to the time of day"
#| fig-width: 4
#| fig-height: 3

# Convert time column to datetime if not already
bus_delay_data["time"] = pd.to_datetime(bus_delay_data["time"], errors="coerce")

# Drop any rows with invalid times
valid_time_data = bus_delay_data.dropna(subset=["time"])

# Extract just the time part for histogram binning
valid_time_data["hour"] = valid_time_data["time"].dt.hour + valid_time_data["time"].dt.minute / 60

# Plot
plt.figure(figsize=(8, 4))
plt.hist(valid_time_data["hour"], bins=24, color="blue", alpha=0.75)

plt.xlabel("Time")
plt.ylabel("Count")
plt.xticks(ticks=range(0, 25, 2), labels=[f"{h}:00" for h in range(0, 25, 2)], rotation=60, ha="right")
plt.grid(axis="y", linestyle=":", linewidth=0.5)
plt.tight_layout()
plt.show()

@@

```{python}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-vstime2
#| fig-cap: "The amount of delays with respect to the time of day"
#| fig-width: 4
#| fig-height: 3

# Ensure the "time" column is interpreted as a datetime
subway_delay_data["time"] = pd.to_datetime(subway_delay_data["time"], errors="coerce")

# Remove rows with invalid times
valid_times = subway_delay_data.dropna(subset=["time"])

# Convert times to fractional hours for histogram binning
valid_times["hour_decimal"] = (
    valid_times["time"].dt.hour + valid_times["time"].dt.minute / 60
)

# Plot histogram
plt.figure(figsize=(8, 4))
plt.hist(valid_times["hour_decimal"], bins=24, color="red", alpha=0.75)

plt.xlabel("Time")
plt.ylabel("Count")
plt.xticks(
    ticks=range(0, 25, 2),
    labels=[f"{h}:00" for h in range(0, 25, 2)],
    rotation=60,
    ha="right"
)
plt.grid(axis="y", linestyle=":", linewidth=0.5)
plt.tight_layout()
plt.show()

@@

```{python}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-avg_num_vsday
#| fig-cap: "The average number of delays occurred on a given day of the week"
#| fig-subcap: ["Bus Delays", "Subway Delays"]
#| layout-ncol: 2

# Define day order
ordered_days = ["Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"]

# Load and format bus data
avg_num_bus_delays_by_day = pd.read_csv("data/summaries/avg_num_bus_delays_by_day.csv")
avg_num_bus_delays_by_day["day"] = pd.Categorical(avg_num_bus_delays_by_day["day"], categories=ordered_days, ordered=True)
avg_num_bus_delays_by_day = avg_num_bus_delays_by_day.sort_values("day")

# Plot bus delays
plt.figure(figsize=(8, 4))
plt.bar(
    avg_num_bus_delays_by_day["day"],
    avg_num_bus_delays_by_day["mean_num_delays"],
    color="skyblue"
)
plt.xlabel("Day of the Week")
plt.ylabel("Mean Number of Delays")
plt.xticks(rotation=60, ha="right")
plt.grid(axis="y", linestyle=":", linewidth=0.5)
plt.tight_layout()
plt.show()

# Load and format subway data
avg_num_subway_delays_by_day = pd.read_csv("data/summaries/avg_num_subway_delays_by_day.csv")
avg_num_subway_delays_by_day["day"] = pd.Categorical(avg_num_subway_delays_by_day["day"], categories=ordered_days, ordered=True)
avg_num_subway_delays_by_day = avg_num_subway_delays_by_day.sort_values("day")

# Plot subway delays
plt.figure(figsize=(8, 4))
plt.bar(
    avg_num_subway_delays_by_day["day"],
    avg_num_subway_delays_by_day["mean_num_delays"],
    color="salmon"
)
plt.xlabel("Day of the Week")
plt.ylabel("Mean Number of Delays")
plt.xticks(rotation=60, ha="right")
plt.grid(axis="y", linestyle=":", linewidth=0.5)
plt.tight_layout()
plt.show()

@@

```{python}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: tbl-totalbus
#| tbl-cap: "The total and average time in minutes wasted on delays"

# Load summarized bus delay data
total_bus_delay_time_by_date = pd.read_csv("data/summaries/total_bus_delay_time_by_date.csv")

# Take 5 random rows with a fixed seed
sample_bus_delays = total_bus_delay_time_by_date.sample(n=5, random_state=302)

# Rename columns to match desired display names
sample_bus_delays.columns = [
    "Date",
    "Day",
    "Total Delay Time",
    "Delay Count",
    "Average Delay Time"
]

# Display the sample table
sample_bus_delays

@@

```{python}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-fullyear
#| fig-cap: "Bus Total and Average Delays in Time (Minutes)"
#| fig-subcap: ["Total Delay Time", "Average Delay Time"]
#| layout-ncol: 2

plt.figure(figsize=(10, 4))
plt.scatter(
    total_bus_delay_time_by_date["date"],
    total_bus_delay_time_by_date["total_delay_time"],
    color="blue", alpha=0.6
)
plt.xlabel("Date")
plt.ylabel("Total Delay Time (Minutes)")
plt.grid(axis="y", linestyle=":", linewidth=0.5)
plt.tight_layout()
plt.show()

plt.figure(figsize=(10, 4))
plt.scatter(
    total_bus_delay_time_by_date["date"],
    total_bus_delay_time_by_date["mean_delay_time"],
    color="green", alpha=0.6
)
plt.xlabel("Date")
plt.ylabel("Total Delay Time (Minutes)")
plt.grid(axis="y", linestyle=":", linewidth=0.5)
plt.tight_layout()
plt.show()

@@

```{python}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-vsincident1
#| fig-cap: "The average number of delays occurred due to incident types"

# Load summarized data
bus_delays_by_incident = pd.read_csv("data/summaries/total_num_bus_delays_by_incident.csv")

# Plot
plt.figure(figsize=(8, 4))
bars = plt.bar(
    bus_delays_by_incident["incident"],
    bus_delays_by_incident["n"],
    color=plt.cm.Set2.colors  # distinct fill colors
)

plt.xlabel("Incident Type")
plt.ylabel("Delay Count")
plt.xticks(rotation=60, ha="right")
plt.grid(axis="y", linestyle=":", linewidth=0.5)

# Add legend manually
incident_labels = bus_delays_by_incident["incident"].tolist()
plt.legend(bars, incident_labels, title="Incident Type", bbox_to_anchor=(1.05, 1), loc="upper left")

plt.tight_layout()
plt.show()

@@

```{python}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-vsincident2
#| fig-cap: "The average number of delays occurred due to incident types"

# Load summarized data
subway_delays_by_incident = pd.read_csv("data/summaries/total_num_subway_delays_by_incident.csv")

# Plot
plt.figure(figsize=(8, 4))
bars = plt.bar(
    subway_delays_by_incident["incident"],
    subway_delays_by_incident["n"],
    color=plt.cm.Set2.colors
)

plt.xlabel("Incident Type")
plt.ylabel("Delay Count")
plt.xticks(rotation=60, ha="right")
plt.grid(axis="y", linestyle=":", linewidth=0.5)

# Add legend
incident_labels = subway_delays_by_incident["incident"].tolist()
plt.legend(bars, incident_labels, title="Incident Type", bbox_to_anchor=(1.05, 1), loc="upper left")

plt.tight_layout()
plt.show()

@@

```{python}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-vsline
#| fig-cap: "The number of delays on each TTC Line"

# Load summarized data
subway_delays_by_line = pd.read_csv("data/summaries/total_num_subway_delays_by_line.csv")

# Plot
plt.figure(figsize=(8, 4))
bars = plt.bar(
    subway_delays_by_line["line"],
    subway_delays_by_line["n"],
    color=plt.cm.Set2.colors  # visually distinct colors
)

plt.xlabel("TTC Line")
plt.ylabel("Number of Delays")
plt.xticks(rotation=60, ha="right")
plt.grid(axis="y", linestyle=":", linewidth=0.5)

# Optional: Add legend
line_labels = subway_delays_by_line["line"].tolist()
plt.legend(bars, line_labels, title="TTC Line", bbox_to_anchor=(1.05, 1), loc="upper left")

plt.tight_layout()
plt.show()
